---
title: "The ARD spec"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_template$set(code = list(comment = "", class.output = "r"))
options(knitr.kable.NA = "")
```

The ARD spec is the study's analyses as a workbook. `tfl_ard_code()`
writes the R program it stands for -- cards / cardx calls, one per
analysis row -- and `tfl_build_ard()` runs that program and saves one
**study ARD** (analysis results data) that every table then reads.

The program is meant to be read. A dataset is read in one line, an
analysis set is one `subset()`, an analysis data is one pipe, an analysis
is one call. It is written the way a programmer would write it, so that
it can be reviewed, copied into a study and finished by hand when needed.

```{r, message = FALSE}
library(tflspec)

study <- file.path(tempdir(), "ard-study")
dir.create(file.path(study, "data"), recursive = TRUE, showWarnings = FALSE)
saveRDS(cards::ADSL, file.path(study, "data", "adsl.rds"))
saveRDS(cards::ADAE, file.path(study, "data", "adae.rds"))

df <- function(...) data.frame(..., stringsAsFactors = FALSE)
```

## The sheets

| Sheet | One row per | Holds |
|---|---|---|
| `study` | key | `id` (the subject key), `output` (where the ARD goes), `source` (R files of the study's own functions) |
| `datasets` | dataset | `dataset`, `level`, `path` (relative to the study folder), `derive` (columns made on reading) |
| `populations` | analysis set | `population_id`, `dataset`, `where`, `derive` |
| `analysis_data` | named data | `data_id`, `label`, `from`, `population_id`, `subjects`, `where`, `add`, `derive`, `keep`, `distinct`, `code` |
| `analyses` | analysis | `output_id`, `analysis_id`, `method`, `data` or `dataset` + `population_id`, `where`, `by`, `strata`, `variables`, `statistics`, `denominator`, `formats`, `args`, `code`, `parent`, `post`, ... |

Every column is described on its header cell's comment in the workbook,
and by `tfl_spec_columns()`:

```{r}
cols <- tfl_spec_columns("analysis_data")
knitr::kable(cols[!is.na(cols$column), c("column", "description")])
```

A new workbook starts from `tfl_ard_spec_template()`; it is read with
`tfl_read_ard_spec()` and written with `tfl_write_ard_spec()`. The
examples below build the sheets in R with `tfl_ard_spec()`, which checks
them exactly as the reader does.

## Datasets and populations

```{r}
base <- list(
  study = df(key = c("id", "output"),
             value = c("USUBJID", "output/ard/ard.rds")),
  datasets = df(dataset = c("ADSL", "ADAE"),
                path = c("data/adsl.rds", "data/adae.rds")),
  populations = df(population_id = "SAF", dataset = "ADSL",
                   where = 'SAFFL == "Y"'))
```

A population is an analysis set: the subjects of `dataset` for which
`where` holds. In the program it is `pop_<population_id>`. An analysis
row may name a `dataset` and a `population_id`: its data are then that
dataset's records of the analysis set's subjects.

```{r, opts.label = "code"}
spec <- tfl_ard_spec(c(base, list(analyses = df(
  output_id = "T-14-1-1", analysis_id = "AGE", method = "continuous",
  population_id = "SAF", by = "TRT01A", variables = "AGE"))))
cat(tfl_ard_code(spec, part = "body"), sep = "\n")
```

## Analysis data

When several analyses read the same data -- treatment-emergent events of
the safety set, say -- the sheet `analysis_data` names it once. Each row
makes one data, in this order:

1. `from` a dataset or an analysis data above it;
2. kept to the subjects of a population (`population_id`) **or** of an
   analysis data above (`subjects`: the numerator kept to the
   denominator's subjects);
3. the records `where` keeps;
4. columns of the population's data added by the subject key (`add`);
5. columns derived (`derive`), and the columns kept (`keep`; with
   `derive`, a `transmute()`);
6. one row per set of values (`distinct`).

`data_id` is the object's name in the program. An analysis names it in
its `data` column instead of `dataset` / `population_id`.

```{r}
adata <- df(
  data_id = c("adsl_saf", "adae_teae", "adae_ser"),
  label = c("Safety set", "TEAEs of the safety set", "Serious TEAEs"),
  from = c("ADSL", "ADAE", "adae_teae"),
  population_id = c("SAF", NA, NA),
  subjects = c(NA, "adsl_saf", NA),
  where = c(NA, 'TRTEMFL == "Y"', 'AESER == "Y"'),
  keep = c(NA, "AEBODSYS | AEDECOD | AESER | TRTA", NA))

analyses <- df(
  output_id = "T-14-3-1",
  analysis_id = c("TEAE", "SER"),
  method = "hierarchical",
  data = c("adae_teae", "adae_ser"),
  by = "TRTA",
  variables = c("AEBODSYS | AEDECOD", "AEDECOD"),
  denominator = "adsl_saf")

spec <- tfl_ard_spec(c(base, list(analysis_data = adata, analyses = analyses)))
```

```{r, opts.label = "code"}
cat(tfl_ard_code(spec, part = "body"), sep = "\n")
```

### What the columns cannot say: `code`

When the columns cannot say how a data is made, `code` is R whose value is
the data. The datasets, the `pop_*` analysis sets and the analysis data
above it are in reach. `from` stays (it says what the data is made from),
and the other columns that make a data are left blank (this is checked):

```{r, opts.label = "code"}
adata2 <- rbind(adata[1:2, ], df(
  data_id = "adae_first", label = "First TEAE of each subject and term",
  from = "adae_teae", population_id = NA, subjects = NA, where = NA,
  keep = NA))
adata2$code <- c(NA, NA, paste(
  "adae_teae <- adae_teae[order(adae_teae$USUBJID, adae_teae$AEDECOD), ]",
  "adae_teae[!duplicated(adae_teae[c(\"USUBJID\", \"AEDECOD\")]), ]",
  sep = "\n"))
spec2 <- tfl_ard_spec(c(base, list(
  analysis_data = adata2,
  analyses = df(output_id = "T-14-3-2", analysis_id = "FIRST",
                method = "categorical", data = "adae_first",
                by = "TRTA", variables = "AEDECOD"))))
cat(tfl_ard_code(spec2, part = "body"), sep = "\n")
```

The program carries it as `<data_id> <- local({ ... })`, in the sheet's
order. The spec keeps the code: the program is written from the spec and
is never edited afterwards.

## Analyses

One row is one cards / cardx call. Both ids become ARD columns, so a
table takes its rows with `tfl_ard_for(ard, output_id)`.

- **`method`**: a keyword -- `tfl_ard_methods()` lists them, each with a
  label a person reads -- or any `pkg::function` of cards / cardx, or a
  function of the study's own (study key `source`).

```{r}
knitr::kable(tfl_ard_methods()[, c("method", "label", "call")])
```

- **`by`, `strata`, `variables`, `statistics`**: `|` between several
  values. `strata` repeats the analysis within each combination present in
  the data (a parameter by visit).
- **`denominator`**: what percentages are of -- `population`, `row`,
  `column`, `cell`, another population, a dataset or an analysis data.
- **`formats`**: `mean=xx.x | p=xx.x%` (the `xx` part says the decimals).
- **`args`**: more arguments, as R (`over_variables = TRUE`). A function
  that takes a formula gets it here: `formula = survival::Surv(AVAL, 1 -
  CNSR) ~ TRTA`.
- **`parent`** and **`post`**: rows run inside `cards::ard_stack()` /
  `ard_strata()` / `ard_pairwise()`, and steps on the ARD after the call
  (`cards::sort_ard_hierarchical()`).

`tfl_ard_functions()` lists every installed cards / cardx `ard_*()`
function by category, and `tfl_ard_args()` says in which column each
argument of one goes:

```{r}
a <- tfl_ard_args("cardx::ard_categorical_ci")
knitr::kable(a[, c("arg", "column", "choices")])
```

### An analysis written as R: `custom`

What no column can say is written as R in `code`, with `method` `custom`.
In the code `data` is the analysis data and `population` the analysis
set's subjects. `tfl_ard_as_custom()` writes an existing row as that code
(the same ARD), as a starting point:

```{r, opts.label = "code"}
cat(tfl_ard_as_custom(spec, "T-14-3-1", "SER")$code, sep = "\n")
```

## Code lists

A table spec's `codelists` sheet (one row a value: `variable`, `value`,
`label`, `order`) gives the order of a variable's values. Handed to the ARD
program, each listed column becomes a factor before any analysis, so the
ARD keeps that order and **counts a value no record has** (`n = 0`).
Rows with a blank `output_id` are the study's; a report's own rows replace
them for that report (`output_id =` one report):

```{r, opts.label = "code"}
codelists <- df(output_id = NA, variable = "SEX", value = c("M", "F"),
                label = c("Male", "Female"), order = 1:2)
spec3 <- tfl_ard_spec(c(base, list(analyses = df(
  output_id = "T-14-1-1", analysis_id = "SEX", method = "categorical",
  population_id = "SAF", by = "TRT01A", variables = "SEX"))))
cat(tfl_ard_code(spec3, part = "body", codelists = codelists), sep = "\n")
```

## Building the ARD

`tfl_build_ard()` runs the program from the study folder. The ARD carries
`output_id`, `analysis_id` and `population_id`; `tfl_ard_conditions()`
lists what went wrong while it was made (cards carries on when one
analysis fails, leaving the message in the ARD).

```{r}
ard <- tfl_build_ard(spec, dir = study, save = FALSE)
table(ard$analysis_id)
tfl_ard_conditions(ard)
```

`tfl_check_ard(ard, spec)` checks an ARD made elsewhere (its shape, and
what the tables read from it); `tfl_write_ard()` writes a copy as JSON,
YAML or XPT. `tfl_ard_spec_hash()` is the fingerprint of an output's
definition, saved with the ARD so a screen can tell which outputs are out
of date.

## The whole program

`tfl_ard_code(spec)` is the whole file: the packages and helpers, the
data, the analyses, `saveRDS()`. Write it to the study's `programs/` and
run it there; tflspec is not needed to run it.

```{r, eval = FALSE}
writeLines(tfl_ard_code(spec), file.path(study, "programs", "ard.R"))
```
