---
title: "CDISC ARS export and import"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(knitr.kable.NA = "")
```

The specs stay the source; CDISC's Analysis Results Standard (ARS v1.0)
is **written from them**, and anyone's ARS can be **read back into them**.
The JSON is the form to exchange; CDISC's Excel template is written for
people to read.

```{r, message = FALSE}
library(tflspec)
df <- function(...) data.frame(..., stringsAsFactors = FALSE)
```

## The specs

An ARD spec, as in the article [The ARD spec](ard-spec.html), plus a
report spec for the outputs' titles. ARS needs one thing the spec does not
guess: each analysis's **`purpose`** (PRIMARY / SECONDARY / EXPLORATORY
OUTCOME MEASURE) is the SAP's decision -- fill the `purpose` column, or
give `tfl_ars(purpose = )` for the rows left blank. `reason` defaults to
SPECIFIED IN SAP.

```{r}
ard_spec <- tfl_ard_spec(list(
  study = df(key = c("id", "study_id"), value = c("USUBJID", "ABC-123")),
  datasets = df(dataset = c("ADSL", "ADAE"),
                path = c("data/adsl.rds", "data/adae.rds")),
  populations = df(population_id = "SAF", dataset = "ADSL",
                   where = 'SAFFL == "Y"'),
  analyses = df(
    output_id = c("T-14-1-1", "T-14-1-1", "T-14-3-1"),
    analysis_id = c("AGE", "SEX", "TEAE"),
    method = c("continuous", "categorical", "hierarchical"),
    dataset = c(NA, NA, "ADAE"),
    population_id = "SAF",
    where = c(NA, NA, 'TRTEMFL == "Y"'),
    by = c("TRT01A", "TRT01A", "TRTA"),
    variables = c("AGE", "SEX", "AEBODSYS | AEDECOD"),
    purpose = "PRIMARY OUTCOME MEASURE")))

report_spec <- tfl_table_spec(
  report = df(output_id = c("T-14-1-1", "T-14-3-1")),
  titles = df(output_id = c("T-14-1-1", "T-14-3-1"), line = 1,
              center = c("Demographic Characteristics",
                         "Treatment-Emergent Adverse Events")))
```

## Writing ARS

```{r}
ars <- tfl_ars(ard_spec, report_spec = report_spec)
tfl_check_ars(ars)      # 0 rows: valid (with jsonvalidate: CDISC's schema too)
```

What ARS has no place for -- and what was added to make it complete --
is listed with the reason, never dropped or added silently:

```{r}
knitr::kable(tfl_ars_unmapped(ars))
```

```{r}
json <- tfl_write_ars_json(ars, file.path(tempdir(), "abc123_ars.json"))
cat(readLines(json, n = 20), sep = "\n")
```

`tfl_write_ars_xlsx(ars, path)` writes CDISC's Excel template, for people
to read.

## Running it with siera

`profile = "siera"` writes ARS that [siera](https://clymbclinical.github.io/siera/)
can run -- still valid CDISC ARS, with an R code template per method.
`tfl_ars_ard()` runs it (siera writes and runs one programme per output)
and gives its ARD, a round trip to compare with the ARD tflspec builds:

```{r, eval = FALSE}
ars_s <- tfl_ars(ard_spec, profile = "siera")
ard <- tfl_ars_ard(ars_s, adam)   # adam: a named list of datasets, or a folder
```

An analysis siera has no template for (a test other than chi-square,
`missing`, `mean_ci`, `custom` ...) is left out and listed by
`tfl_ars_unmapped()`.

## Reading anyone's ARS

`tfl_read_ars_json()` reads an ARS JSON -- one written elsewhere, by
another tool -- and `tfl_ars_to_specs()` turns it into specs: `$ard`,
`$report` (and with `table = TRUE`, `$table`: the groupings' levels).
ARS has no ADaM paths, so `datasets$path` is filled in before the ARD spec
runs; what has no place in the specs is in `attr(, "unmapped")`.

```{r}
sp <- tfl_ars_to_specs(tfl_read_ars_json(json))
sp$ard$analyses[, c("output_id", "analysis_id", "method", "by", "variables")]
sp$report$titles[, c("output_id", "line", "center")]
```

There is no way back from an ARD alone (it has no data paths, populations
or conditions); an existing study's analyses come in through ARS.
