---
title: "Listings"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_template$set(code = list(comment = "", class.output = "r"))
```

A listing is two sheets keyed by `output_id`: `listings`, one row a
listing (its dataset, the records it keeps, the order, rows a page), and
`listing_cols`, one row a printed column. `tfl_listing_code()` writes the
program -- read the data, keep, sort, lay out -- and `tfl_listing()` makes
the pages from data in hand. The layout is rtfreporter's
`listing_spec()` / `as_rtftables()`.

```{r, message = FALSE}
library(tflspec)
library(rtfreporter)
```

## The sheets

| Sheet | Columns |
|---|---|
| `listings` | `output_id`, `type` (an rtfreporter listing type; blank: `multiline`), `dataset`, `where` (R), `sort` (`-` for descending), `max_rows`, `blank_row`, `wrap` |
| `listing_cols` | `output_id`, `vars` (`|` stacks variables in one column), `label`, `width` (characters), `sep`, `align`, `collapse_repeats` |

```{r}
spec <- tfl_listing_spec(
  listings = data.frame(output_id = "L-16-2-7", dataset = "ADAE",
                        where = 'AESER == "Y"',
                        sort = "USUBJID | ASTDY", max_rows = "15"),
  listing_cols = data.frame(
    output_id = "L-16-2-7",
    vars  = c("USUBJID", "TRTA", "AEBODSYS | AEDECOD", "ASTDY", "AESEV"),
    label = c("Subject", "Treatment", "System organ class /\nPreferred term",
              "Study day", "Severity"),
    width = c("14", "22", "40", "8", "10"),
    collapse_repeats = c("TRUE", "TRUE", NA, NA, NA)))
```

A workbook is read with `tfl_read_listing_spec()` and written with
`tfl_write_listing_spec()`; `tfl_write_listing_spec(tfl_listing_spec(),
path)` gives an empty one to fill in.

## The program

The data catalog -- the `datasets` sheet of the ARD spec: each dataset's
`path` and derived columns -- says how the data is read, so the listing's
program reads it the same way the ARD program does:

```{r, opts.label = "code"}
catalog <- data.frame(dataset = "ADAE", path = "data/adae.rds")
cat(tfl_listing_code(spec, "L-16-2-7", datasets = catalog), sep = "\n")
```

What the sheets cannot say about the data -- a derived column, a join --
is `rework`: code run on `data` before it is sorted, written into the
program by the spec, not added afterwards by hand:

```{r, opts.label = "code"}
cat(tfl_listing_code(spec, "L-16-2-7", datasets = catalog,
                     rework = "data$AEDECOD <- tools::toTitleCase(tolower(data$AEDECOD))"),
    sep = "\n")
```

The report around it is the report spec's, as for a table:
`tfl_report_code(report_spec, "L-16-2-7", content = "content")`.

## The pages

`tfl_listing()` does what the program does, on data in hand. The data
comes first, so it pipes:

```{r}
pages <- cards::ADAE |> tfl_listing(spec, "L-16-2-7")
length(pages)
class(pages[[1]])
```

The pages go to `tfl_report(report_spec, content = pages)` or
`rtf_tables()`.

## A listing written in code, back to a spec

`tfl_as_listing_spec()` writes a listing coded with rtfreporter
(`listing_spec()`, or a `plan_listing()` plan) back as a listing spec,
listing what the sheets cannot carry (`attr(, "not_converted")`):

```{r}
lst <- listing_spec(list(
  listing_col("USUBJID", label = "Subject", width = 14),
  listing_col(c("AEDECOD", "AESEV"), label = "Term / Severity", width = 30)))
back <- suppressMessages(tfl_as_listing_spec(lst, "L-16-2-8", dataset = "ADAE"))
back$listing_cols[, c("vars", "label", "width")]
```
