---
title: "The table spec"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_template$set(code = list(comment = "", class.output = "r"))
options(knitr.kable.NA = "")
```

The table spec says how an ARD becomes a table: which keys go across and
down, the label and order of each variable, the template and digits of
each cell, the pages, the column widths, the header. It stands for an
[rtfreporter](https://github.com/ichirio/rtfreporter) **plan**:
`table_plan()` with the roles, and one `plan_*()` verb for each thing the
other sheets say.

- `tfl_table_code(spec)` writes that plan as code -- the program a study
  keeps, which no longer reads the workbook;
- `tfl_table_plan(data, spec)` builds the same plan from an ARD.

```{r, message = FALSE}
library(tflspec)
library(rtfreporter)
```

## The sheets

One sheet per grain, so each fact is written once. Column names are
rtfreporter's argument names: a `layout` column is the verb's prefix and
its argument (`pages_max_rows` is `plan_paginate_rows(max_rows = )`).

| Sheet | One row per | Goes to |
|---|---|---|
| `study` | study fact | `rounding`, `output_path`, `program_dir` |
| `tables` | table | `table_plan(cols, rows, label)`, `plan_cells()`, `plan_sort()` |
| `variables` | variable | `plan_labels()`, `plan_levels()` |
| `codelists` | value of a variable | `plan_labels()`, `plan_levels()` (and the ARD's factors) |
| `cells` | line of a cell | `plan_cells()`, `plan_digits()` |
| `layout` | table | `plan_paginate_rows()`, `plan_row_group()`, `plan_blanks()`, `plan_stub()`, `plan_paginate_cols()` |
| `columns` | printed column | `plan_columns()`, `plan_hide()` |
| `style` | table | `plan_style()` |
| `cell_styles` | styling | `plan_cell_style()` |
| `col_header` | header cell | `plan_col_header()` |

The **`output_id` rule**: a row with a blank `output_id` is the study's
default; a report's own row replaces it, per sheet (and, on the keyed
sheets, per key). Every column is described on its header cell's comment
in the workbook and by `tfl_spec_columns()`.

## The example workbooks

The package comes with the specs of five tables -- demographics (`DM`),
adverse events by SOC and PT (`AE`), response by subgroup (`ORR`), a
lab shift table (`LB`) and PK concentrations (`PK`) -- each in its own
workbook, and all five in `study.xlsx`:

```{r}
ex <- system.file("extdata", "ard-spec", package = "tflspec")
spec <- tfl_read_table_spec(file.path(ex, "study.xlsx"))
spec$tables[, c("output_id", "cols", "rows", "label", "sort")]
```

The AE table: treatment groups across, SOC as row groups, PT as rows,
sorted by frequency.

```{r, opts.label = "code"}
cat(tfl_table_code(spec, "AE", pipe = "|>"), sep = "\n")
```

## Cells

`cells` rows are templates: `{mean} ({sd})`, `{n} ({p:.1f%})`. A row with
a blank `variable` is the table's default template. Rows with the same
variable / context / row form one chain tried in order, and `when` is an
R guard (`n == 0`). In a table of one statistic per row
(`tables$stats = rows`), a row with no template gives one statistic's
digits.

```{r}
spec$cells[spec$cells$output_id %in% "ORR",
           c("variable", "row", "when", "template")]
```

## From an ARD: a template to start from

`tfl_table_spec_template()` walks an ARD and writes the `tables`,
`variables` and `cells` sheets -- the variables with their levels, a
template for each kind of row -- to be edited:

```{r}
ard <- cards::ard_stack(
  cards::ADSL, .by = TRT01A,
  cards::ard_summary(variables = AGE),
  cards::ard_tabulate(variables = AGEGR1))
tmpl <- tfl_table_spec_template(ard, cols = "TRT01A", output_id = "T-DM")
tmpl$variables
tmpl$cells[, c("variable", "row", "template")]
```

Write it with `tfl_write_table_spec(tmpl, "spec/tables.xlsx")`, edit it in
Excel, read it back with `tfl_read_table_spec()`.

## A plan, and one report's own change

`tfl_table_plan()` returns a plan, so any rtfreporter verb piped after it
wins over the workbook ("last wins") -- one report departs from the study's
workbook in one line, and the workbook is not touched:

```{r}
plan <- tfl_table_plan(normalize_ard(ard), tmpl) |>
  plan_cells(notes = FALSE) |>
  plan_labels(AGE = "Age (years)", AGEGR1 = "Age group, n (%)")
knitr::kable(plan_apply(plan, stage = "table"))
```

## A plan written in code, back to a workbook

`tfl_as_table_spec()` goes the other way: a plan written in code becomes
a workbook. It compares the RTF the two make (byte by byte) and lists
what a sheet cannot carry (`attr(, "not_converted")`) -- a look computed
row by row, a step after the pages are made -- which stays in code, after
the plan:

```{r}
coded <- table_plan(normalize_ard(ard), cols = "TRT01A",
                    rows = c(group = "variable")) |>
  plan_cells(continuous = c("Mean (SD)" = "{mean} ({sd})"),
             categorical = "{n} ({p:.1f%})", notes = FALSE) |>
  plan_paginate_rows(max_rows = 30)
back <- tfl_as_table_spec(coded, output_id = "T-DM")
back$cells[, c("variable", "row", "template")]
back$layout[, "pages_max_rows", drop = FALSE]
```

## Reports and several workbooks

A table spec is usually read together with its report spec (the page,
titles, footnotes): `tfl_read_report_spec(c("report.xlsx",
"tables.xlsx"), output_id = )` reads both, and one object serves
`tfl_table_plan()` and `tfl_report()` alike. A sheet may be in any of the
workbooks given, but in only one. See [Reports](reports.html).
