---
title: "Get started with tflspec"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# generated code is shown as code
knitr::opts_template$set(code = list(comment = "", class.output = "r"))
options(knitr.kable.NA = "")
```

tflspec reads the specifications of a study's tables, listings and figures
(Excel workbooks, YAML for figures) and writes the **R program** each one
stands for. The program is the deliverable: short, readable, written the
way a person would write it, and needing only cards / cardx and
[rtfreporter](https://github.com/ichirio/rtfreporter) to run -- not
tflspec. People copy it into the study, read it, and finish it.

```text
ADaM ──(ARD spec)──> ARD ──(table spec)──> plan ──(report spec)──> RTF
          tfl_ard_code()      tfl_table_code()       tfl_report_code()
          tfl_build_ard()     tfl_table_plan()       tfl_report()
```

Each spec can be **written out as code** (the `*_code()` functions) or
**run** (`tfl_build_ard()`, `tfl_table_plan()`, `tfl_report()`). Both come
from the same list of steps, so the object and the program cannot
disagree.

This article walks the whole flow for one table, a demographics table
(`DM`), using the workbooks that come with the package and the ADaM data
that come with cards.

```{r setup, message = FALSE}
library(tflspec)
library(rtfreporter)
```

## A study folder

Every path in a spec is relative to the study folder, and the generated
programs run from it. Ours holds two ADaM datasets:

```{r}
study <- file.path(tempdir(), "study")
dir.create(file.path(study, "data"), recursive = TRUE, showWarnings = FALSE)
saveRDS(cards::ADSL, file.path(study, "data", "adsl.rds"))
saveRDS(cards::ADAE, file.path(study, "data", "adae.rds"))
```

## 1. The ARD spec: what is analysed

The ARD spec is a workbook of a few sheets: `study` (keys such as the
subject key), `datasets` (what is read, and columns derived on reading),
`populations` (the analysis sets), `analysis_data` (named data the analyses
read, each a report's own; optional) and `analyses`, **one cards / cardx
call per row**.
A study normally keeps it as an `.xlsx`; here it is written from R so the
whole article runs on its own.

```{r}
df <- function(...) data.frame(..., stringsAsFactors = FALSE)

ard_spec <- tfl_ard_spec(list(
  study = df(key = c("id", "output"),
             value = c("USUBJID", "output/ard/ard.rds")),
  datasets = df(dataset = c("ADSL", "ADAE"),
                path = c("data/adsl.rds", "data/adae.rds"),
                derive = c("HTBL = HEIGHTBL", NA)),
  populations = df(population_id = "SAF", dataset = "ADSL",
                   where = 'SAFFL == "Y"'),
  analyses = df(output_id = "DM",
                analysis_id = c("BIGN", "CONT", "CAT"),
                method = c("total_n", "continuous", "categorical"),
                population_id = "SAF",
                by = c(NA, "TRT01P", "TRT01P"),
                variables = c(NA, "AGE | HTBL", "AGEGR1 | SEX | ETHNIC"))))

path <- tfl_write_ard_spec(ard_spec, file.path(study, "ard_spec.xlsx"))
ard_spec <- tfl_read_ard_spec(path)
ard_spec$analyses[, c("output_id", "analysis_id", "method", "by", "variables")]
```

`method` is a keyword (`tfl_ard_methods()`), any `pkg::function` of cards
or cardx, or a function of the study's own. The article
[The ARD spec](ard-spec.html) goes through every sheet.

### The ARD program

`tfl_ard_code()` writes the program. It reads the data, makes the analysis
set, and calls cards once per analysis:

```{r, opts.label = "code"}
cat(tfl_ard_code(ard_spec, part = "body"), sep = "\n")
```

`part = "body"` leaves out the setup every program starts with (the
packages, the helpers that tag each result with its `output_id` /
`analysis_id`); `tfl_ard_code(ard_spec)` writes the whole file, ready to
save as `programs/ard.R`.

### The ARD

`tfl_build_ard()` runs that same program from the study folder and saves
the study ARD where the `study` key `output` says:

```{r}
ard <- tfl_build_ard(ard_spec, dir = study)
ard[1:6, c("output_id", "analysis_id", "variable", "stat_name", "stat")]
```

## 2. The table spec: how the ARD becomes a table

The table spec says which keys go across (`cols`) and down (`rows`), the
labels and order of each variable, the cell templates and digits, the
pages, the column header. The package's example `DM.xlsx` defines this
table:

```{r}
ex <- system.file("extdata", "ard-spec", package = "tflspec")
spec <- tfl_read_report_spec(file.path(ex, c("report.xlsx", "DM.xlsx")),
                             output_id = "DM")
spec$cells[, c("variable", "row", "template", "digits")]
```

`tfl_read_report_spec()` reads the table sheets and the report sheets
together (one object serves both); `tfl_read_table_spec()` reads the table
sheets alone.

### The table program

`tfl_table_code()` writes the rtfreporter plan: the roles in
`table_plan()`, then one `plan_*()` verb for each thing the sheets say.

```{r, opts.label = "code"}
cat(tfl_table_code(spec, pipe = "|>"), sep = "\n")
```

### The table

`tfl_table_plan()` builds that same plan from an ARD. Being a plan, a verb
piped after it still wins -- one report's own change, in one line, without
touching the workbook:

```{r}
data <- normalize_ard(tfl_ard_for(ard, "DM"))
plan <- tfl_table_plan(data, spec) |>
  plan_cells(notes = FALSE)
```

The table it makes (`plan_apply(stage = "table")` stops before the pages):

```{r}
knitr::kable(plan_apply(plan, stage = "table"))
```

## 3. The report spec: the document around the table

The report spec dresses the page: paper and margins, the running header
and footer, titles and footnotes, tokens such as `{PAGE}` and
`{PROGRAM}`. The study's header is written once, as default rows (a blank
`output_id`), and each report adds its own lines:

```{r}
spec$header[, c("line", "left", "center", "right")]
```

### The report program

```{r, opts.label = "code"}
cat(tfl_report_code(spec, content = "plan"), sep = "\n")
```

### The RTF

```{r}
doc <- tfl_report(spec, content = plan)
rtf <- file.path(study, tfl_report_path(spec))
dir.create(dirname(rtf), showWarnings = FALSE)
generate_rtfreport(doc, rtf, overwrite = TRUE)
file.exists(rtf)
```

## The program a study keeps

Put together, the three pieces are one short program per output that a
person can read top to bottom:

```{r, eval = FALSE}
library(rtfreporter)
# 1 DATA    the ARD (programs/ard.R, written by tfl_ard_code())
ard  <- readRDS("output/ard/ard.rds")
data <- normalize_ard(ard[ard$output_id == "DM", ])
# 2 TABLE   tfl_table_code()
plan <- table_plan(data, cols = "TRT01P", rows = c(group = "variable")) |>
  plan_labels(AGE = "Age (years) [a]") |>
  ...
# 3 REPORT  tfl_report_code()
doc <- rtf_document(program = "DM")
doc <- rtf_tables(doc, plan)
generate_rtfreport(doc, "output/DM.rtf", overwrite = TRUE)
```

## When the spec cannot say it

A spec says the common things in columns. What it cannot express is
written as R **in the spec**, in its code columns, so the program is still
written from the spec and never edited by hand afterwards:

- `analysis_data$code`: R that makes an analysis data;
- `analyses$code` with method `custom`: R that makes an analysis's ARD
  (`tfl_ard_as_custom()` writes an existing row as such code, to start
  from);
- `args`, `derive`, `where`: R expressions where a value is code;
- a figure design's `data_code` and `layer_code` steps, and `!r` values.

And what is truly one report's own -- a cell style computed row by row, a
step after the pages are made -- is one more verb piped after the plan.

## Where next

- [The ARD spec](ard-spec.html): datasets, populations, a report's
  analysis data (`subjects`, `keep`, `code`), its code lists, the
  analyses.
- [The table spec](table-spec.html): the table sheets, numbers or the
  ARD's text, the decimals written once, a report's code lists and the
  variables' labels, and a plan written in code back to a workbook.
- [Reports](reports.html): headers, footers, titles, footnotes and tokens
  written once for the study, each report's own tokens, the company's
  font and the setup code the report programs share.
- [Listings](listings.html), [Figures](figures.html) and
  [CDISC ARS](ars.html).
- Working with an AI assistant? Attach `tflspec_ai_manual()`, the manual
  of the installed version.
